---
title: "LULC"
author: "Sophie Spelsberg"
date: "2024-01-08"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

# Land Cover Classification

Here, I will train a machine learning algorithm to model land cover
classes from MODIS data, based on the land cover classification model
implemented by Hufken (2023) in *A Handful of Pixels*. As training and
test data set, we will use the data sets provided there. Original
validation data stems from Fritz et al. (2017) and can be downloaded via
Zenodo. As predictor data we use MODIS data from the MCD43A4 data
product (bands 1-7). This can be downloaded with the NASA AppEEARS API
via the {appears} package.

There are various options to improve the model performance in comparison
to the original model. The implementation in *A Handful of Pixels* is a
simplified version of the official MODIS land use and land cover (LULC)
product (MCD12Q1 LULC), which shows a much better model performance than
the simplified version (Friedl et al. 2010). Therefore, some
improvements can be derived from their implementation of the model.

1)  **tune more hyper-parameters:** this can increase the model
    performance significantly, but comes at higher computational costs.
2)  **increase the number of trees:** can also lead to a better model
    performance, also at higher computational costs.
3)  **use monthly means for the predictor variables and include annual
    mean, minimum and maximum values in the predictor set**: this seems
    counterintuitive, as it decreases the temporal resolution of the
    MCD43A4 data. However, it might represent the seasonal changes in
    the data better than using daily values as individual predictors.
    The annual minimum, maximum and mean values also add information
    about the annual course of the data that is not reflected in the
    daily data set. These parameters are also used by Friedl et al.
    (2010). Moreover, these changes reduce the computation time, which
    allows the rest of the model to be more complex.
4)  **add surface temperature as a predictor**: temperature has a strong
    influence on land cover - in alpine regions it determines the
    boundary between forest, shrubs and grassland, and is a crucial
    factor whether an area is suitable for agricultural use. Therefore,
    it makes sense to include additional temperature variables as
    predictors (e.g. min, max, mean temperature, monthly means) like the
    official MODIS LULC product does.
5)  **increase the number of validation sites**: a good spatial
    distribution and high number of training data points can increase
    the model performance. Also, as our model is only applied to
    Switzerland, it might make sense to restrict the validation data to
    Switzerland and surrounding areas.
6)  **validate the results over multiple years**: Friedl et al. (2010)
    use this technique to control if the classification of a pixel
    changed between two years without actual changes in land cover. This
    helps to identify classification errors.

In the following, I will implement the first three ideas: I will tune
more hyper-parameters, increase the number of trees and I will use
monthly instead of daily predictor data and include annual mean, minimum
and maximum values. Thus, I will add parts to the model training
workflow as well as an additional step to the data wrangling.

```{r libraries, include = FALSE}
use_pkgs <-  c("dplyr", "tidyr", "readr", "lubridate", "stringr", "purrr",
              "ggplot2", "tidyverse", "visdat", "terra", "hexbin", "jsonlite",
              "MODISTools", "forcats", "yardstick", "recipes", "caret",
              "broom", "skimr", "cowplot", "scico", "hwsdr", "usethis",
              "renv", "modelr", "rmarkdown", "rpart", "RColorBrewer",
              "rpart.plot", "ranger", "sessioninfo", "skimr", "ncdf4", "pdp", 
              "vip")

# read all into library
lapply(use_pkgs, library, character.only=TRUE)

# important for this model:
library(parsnip) # tidy machine learning
library(workflows) # manage workflows
library(rsample) # split into folds
library(tune)
library(dials) # (hyper) parameter sampling schemes
library(xgboost) # for boosted regression tree
```

## Data wrangling

After loading the data sets provided in *A Handful of Pixels,* we can
begin with the data wrangling. In this case, we want to change the
predictor data from daily to monthly data to represent the annual cycle
better.

```{r load train and test data}
# wide format 
# cols: pixelID, LC1, lat, lon, bands at given dates
train <- readRDS("../data/training_data.rds")
test <- readRDS("../data/test_data.rds")
```

To do so, it is important to understand the structure of the data. It
comes in a wide format, so every row represents one validation location.
The first four columns of the train dataframe are the ID of the Pixel,
its land cover classification from Fritz et al. (2017), its longitude
and its latitude. Afterwards, each column contains the measurement for
one band and for a single day. The test dataframe has the same
structure, but without the first four columns.

The simplified model uses each day as a single predictor. To generate
monthly data per band from the train and test dataframe, I created a
function that takes the original dataframe. To extract the information
on date and band from the name of the columns, it first converts the
dataframe into a longer format with only one column for all reflectance
values and one column with all former column names. Afterwards, they can
be read via the {stringr} package and converted into timestamps with
{lubridate} and numeric information on the band.

From this tidy data structure, annual mean, minimum and maximum values
can be computed per band. Additionally, monthly means per band can be
calculated. By converting both dataframes back into a wide format and
joining them, the result is a data format like the original, where each
row corresponds to one pixel - but with monthly data instead of daily
and additional annual mean, min and max values.

However, to retreive data that the boosted regression tree model can
process, one additional step is necessary. As parts of the data are
missing, some of the built in functions (e.g. min()) return warnings as
they generate NaN and Infinite values. Those have to be converted to NA
values as the last step of data wrangling.

To additionally preserve the information on PixelID and land cover
classification in the train dataframe, the function takes information on
whether it is the train or the test data set and performs some
operations only on the train dataframe.

```{r data wrangling: change to monthly data}

# change depending on train and test
get_monthly_predictors <- function(data, test_data=FALSE){
  
  data <- data[,1:2566] |>  # cut after reflection bands
    mutate(own_ID = c(1:nrow(data))) |>  # add ID to reorder the df in the end
    
    # convert to a longer format (only one column with reflectance values)
    tidyr::pivot_longer(
      cols = starts_with("MCD43A4_061_Nadir_Reflectance"),
      names_to = "band_date",
      values_to = "val"
      ) |>
    
    # get date and band information from band_date string
    # turn last 10 characters into datetime object
    mutate(date = lubridate::ymd(stringr::str_sub(band_date, -10)),
           month = lubridate::month(date),
           band = as.numeric(stringr::str_sub(band_date, -12, -12))
           )

  data_annual <- data |>
    
    # compute annual mean, min and max values
    group_by(own_ID, band) |>
    summarise(
      mean = mean(val, na.rm=TRUE),
      min = min(val, na.rm=TRUE),
      max = max(val, na.rm=TRUE)
      ) |>
    ungroup() |>
    
    # change format
    tidyr::pivot_longer(
      cols = c(mean, min, max),
      names_to = "type",
      values_to = "val"
      ) |> 
    mutate(name = paste0("band", band, "_", type)) |>
    dplyr::select(!c(band, type)) |>
    tidyr::pivot_wider(
      names_from = name,
      values_from = val
      )
  
  # preserve information on PixelID and LC1 in the train data
  if (test_data) {
    
      data <- data |> 
    
        # compute monthly means
        group_by(own_ID, band, month) |>
        summarise(
          mean = mean(val, na.rm=TRUE)
          )    
    
  } else {
    
      data <- data |> 
    
        # compute monthly means
        group_by(own_ID, band, month) |>
        summarise(
          LC1 = mean(LC1),               # additional column for train
          pixelID = mean(pixelID),       # additional column for train
          mean = mean(val, na.rm=TRUE)
          )
  }  

  data <- data |>
    ungroup() |>
    mutate(band_date = paste0("band", band, "_month", month)) |>
    dplyr::select(!c(band,month)) |> # drop band, month column
    
    # turn into wider format
    tidyr::pivot_wider(
      values_from = mean,
      names_from = band_date
      ) 
  
  # add monthly and annual data and turn NaN and Inf to NA
  data <- left_join(data, data_annual) |>
    mutate_all(
      ~ifelse(
        (is.nan(.) | is.infinite(.)), 
        NA, 
        .)
      )
  
  return(data)
}


# convert train and test set into new format
train_monthly <- get_monthly_predictors(train) 
test_monthly <- get_monthly_predictors(test, test_data=TRUE)

```

## Model training

To train the model in the next step, it is necessary to first split the
data in an independent train and test set to evaluate the model
performance later on (as the provided test set does not contain
information on the land cover, it cannot be used for this purpose).

```{r split data}

# split data along land cover classes
monthly_split <- train_monthly |>
  dplyr::select(LC1, contains("band")) |>
  rsample::initial_split(strata = LC1, prop = 0.8)

# select training and testing data
train2 <- rsample::training(monthly_split)
test2 <- rsample::testing(monthly_split)
```

Afterwards, the model structure is specified via the {parsnip} package.
It provides an interface for the {xboost} package - we use a boosted
regression / classification tree. In this step, I implement the second
improvement: tuning additional hyper-parameters. As additional
parameter, I chose the learn rate, which was not tuned in the original
implementation. This increases the computation time significantly.
However, as the number of predictors has been reduced extremely (from
2608 to 105), the computation time is much lower than when applying this
step to the original model. In theory, it is possible to tune various
further parameters for a boosted regression tree (e.g. mtry, the number
of predictors included per tree). However, to keep the computation time
for this model low, I will only tune one further parameter in this
example.

As a next step, the workflow is defined with the {workflow} package.

```{r define model & workflow}

# specify model structure and model - final model
model_settings <- parsnip::boost_tree(
  trees = 500, 
  min_n = tune(),
  tree_depth = tune(),
  learn_rate = tune() # additionally tuned
  ) |>
  set_engine("xgboost") |>
  set_mode("classification")

# create workflow compatible with {tune} package
xgb_workflow <- workflows::workflow() |>
  add_formula(as.factor(LC1) ~ .) |>
  add_model(model_settings)
```

Then, additional settings are made for tuning the hyper-parameters:

```{r hyperparameter settings}

# use dials package for hyper parameter settings
hp_settings <- dials::grid_latin_hypercube(
  tune::extract_parameter_set_dials(xgb_workflow),
  size = 3
  )
```

Afterwards, the actual model can be calibrated as in *A Handful of
Pixels*, including a three-fold cross-validation. The best model is
selected in the end and trained on the data. The tuned hyper-parameters
for a tree number of 500 are min_n = 35 , tree_depth = 2 and learn_rate
= 0.02.

```{r model calibration}

set.seed(3) # for reproducibility

# create folds (divide into different cross-validation training datasets)
folds <- rsample::vfold_cv(train2, v = 3)

# optimize the model (hyper) parameters and use:
# 1. the workflow (i.e. model)
# 2. the cross-validation across training data
# 3. the (hyper) parameter specifications
# all data are saved for evaluation
xgb_results <- tune::tune_grid(
  xgb_workflow,
  resamples = folds,
  grid = hp_settings,
  control = tune::control_grid(save_pred = TRUE)
)

# select the best model based upon RMSE
xgb_best <- tune::select_best(
  xgb_results,
  metric = "roc_auc"
  )

# cook up a model using finalize_workflow
# which takes workflow (model) specifications
# and combines it with optimal model
# parameters into a model workflow
xgb_best_hp <- tune::finalize_workflow(
  xgb_workflow,
  xgb_best
)

# train a final model with tuned hyper-parameters
xgb_best_model <- fit(xgb_best_hp, train2)

# print fitted hyper-parameters
print(xgb_best_hp$fit$actions$model$spec)

```

## Model evaluation

To test the performance of the model, we can first predict the land
cover classes of our test set and than access the accuracy of the
predictions. This can be visualized with the confusion matrix provided
by the caret package:

```{r evaluate my own model}

# run the model on the test2 split
test_results <- predict(xgb_best_model, test2)

# overview of metrics for model evaluation
conf_matrix <- caret::confusionMatrix(
  reference = as.factor(test2$LC1),
  data = as.factor(test_results$.pred_class)
  )

# results
print(conf_matrix$overall[1:2])
print(conf_matrix$table)


```

The accuracy of the model from *A Handful of Pixels* is 0.42, the
accuracy of the final model is 0.47 (the result differs slightly betwen
the runs). To access the influence of the single improvements on the
model performance, I also evaluated the model

1) with only monthly predictor data (Accuracy: 0.44)

2) with additional hyper-parameter tuning (Accuracy: 0.44)

3) with hyper-parameter tuning and an additionally increased number of
trees (Accuracy: 0.47)

It can be seen that the additional tuning of the learn rate has almost
no influence on the model performance, while the higher number of trees
and the monthly predictor data do make a difference. Interesting is
especially the effect of monthly predictor data. By reducing the number
of predictors and decreasing the temporal resolution, the model reduces
its computational time to a fraction of the original time while
increasing the accuracy by 0.02. This highlights the importance of data
wrangling and the choice of the right predictors. It also shows that
choosing daily data as individual predictors might not be the best
choice when the actual predictor should be an annual cycle.

# References

Friedl, M. et al. 2010. MODIS Collection 5 global land cover: Algorithm
refinements and characterization of new datasets. Remote Sensing of
Environment, 114, 1, 168-182.
[doi.org/10.1016/j.rse.2009.08.016](https://doi.org/10.1016/j.rse.2009.08.016 "Persistent link using digital object identifier")

Fritz, S. et al. 2017. A global dataset of crowdsourced land cover and
land use reference data. Scientific Data, 4, 170075.
[doi.org/10.1038/sdata.2017.75](https://doi.org/10.1038/sdata.2017.75)

Hufkens, K. 2023. A Handful of Pixels.
<https://geco-bern.github.io/handfull_of_pixels/>
